---
title: "Getting Started with rDeckgl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rDeckgl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**rDeckgl** provides R bindings for [deck.gl](https://deck.gl), a WebGL framework for visualizing large datasets. Key features:

- **Zero-copy data transfer**: DuckDB exports Arrow IPC streams directly to the browser
- **Efficient rendering**: WebGL-powered visualization handles millions of points
- **GeoArrow support**: Native spatial data types (points, polygons, lines) with automatic WKB conversion
- **Flexible specs**: Define visualizations using R lists, JSON, or YAML

## Architecture

The package uses a custom JavaScript bundle (`deckgl-bundle.js`) that combines:
- `@deck.gl/core` and layer packages (standard + geo + GeoArrow)
- `apache-arrow` for client-side Arrow decoding
- Custom WKB→GeoArrow polygon converter (DuckDB exports WKB, deck.gl expects nested lists)

Data flow:
1. R writes data frames to in-memory DuckDB (or data already exists in DuckDB)
2. SQL queries → `COPY ... (FORMAT ARROW)` exports binary Arrow IPC
3. Browser fetches Arrow, decodes, renders via WebGL/deck.gl via geoarrow/deckgl-layers

**Shiny vs Static**: 
- **Static** (htmlwidgets): Arrow data embedded as Base64 in HTML (~33% size overhead)
- **Shiny**: Arrow served as binary via `session$registerDataObj` (true zero-copy, no encoding)

Use Shiny when data is large (>5MB) or needs reactivity. Use static for reports/documents.

## Installation

```{r install}
remotes::install_github("TiRizvanov/rDeckgl")
```

## Quick Example

Visualize 1000 random points:

```{r scatterplot}
library(rDeckgl)
library(duckdb)
library(DBI)

con <- dbConnect(duckdb())
dbExecute(con, "INSTALL spatial; LOAD spatial;")

dbExecute(con, "
  CREATE TABLE points AS
  SELECT
    ST_Point(
      -122.4 + (random() - 0.5) * 0.2,
      37.76 + (random() - 0.5) * 0.2
    )::POINT_2D as geometry
  FROM range(1000)
")

df <- dbGetQuery(con, "SELECT * FROM points")
dbDisconnect(con)

spec <- list(
  initialViewState = list(longitude = -122.4, latitude = 37.76, zoom = 11),
  layers = list(
    list(
      `@@type` = "GeoArrowScatterplotLayer",
      data = list(type = "duckdb", query = "SELECT geometry FROM points"),
      getPosition = "@@col:geometry",
      getFillColor = c(0, 150, 255),
      getRadius = 50
    )
  )
)

deckgl(spec, data = list(points = df))
```

## Key Functionality

### DuckDB Integration

Use SQL to transform data before visualization:

```{r duckdb_query}
library(rDeckgl)
library(duckdb)
library(DBI)

# Prepare data with spatial points
con <- dbConnect(duckdb())
dbExecute(con, "INSTALL spatial; LOAD spatial;")

dbExecute(con, "
  CREATE TABLE sales AS
  SELECT
    i as id,
    random() * 100 as value,
    ST_Point(-122.4 + random() * 0.3, 37.76 + random() * 0.2)::POINT_2D as geometry
  FROM range(500) t(i)
")

# Use SQL to filter and scale - only high-value points
df <- dbGetQuery(con, "SELECT * FROM sales WHERE value > 50")
dbDisconnect(con)

spec <- list(
  initialViewState = list(longitude = -122.4, latitude = 37.86, zoom = 11),
  layers = list(
    list(
      `@@type` = "GeoArrowScatterplotLayer",
      data = list(
        type = "duckdb",
        query = "SELECT geometry, value FROM sales WHERE value > 50"
      ),
      getPosition = "@@col:geometry",
      getFillColor = c(255, 50, 50),
      getRadius = 100
    )
  )
)

deckgl(spec, data = list(sales = df))
```

### Polygon Visualization

Render polygons using native `POLYGON_2D` geometry:

```{r polygons}
library(rDeckgl)
library(duckdb)
library(DBI)

con <- dbConnect(duckdb())
dbExecute(con, "INSTALL spatial; LOAD spatial;")

# Create square polygons using native POLYGON_2D type
dbExecute(con, "
  CREATE TABLE zones AS
  SELECT
    ST_MakePolygon(
      ST_MakeLine([
        ST_Point(x, y),
        ST_Point(x + 0.008, y),
        ST_Point(x + 0.008, y + 0.008),
        ST_Point(x, y + 0.008),
        ST_Point(x, y)
      ])
    )::POLYGON_2D as geometry,
    'Zone_' || i as id
  FROM (
    SELECT
      i,
      -122.45 + (i % 5) * 0.015 as x,
      37.75 + floor(i / 5) * 0.01 as y
    FROM range(10) t(i)
  )
")

df <- dbGetQuery(con, "SELECT * FROM zones")
dbDisconnect(con)

spec <- list(
  initialViewState = list(longitude = -122.4, latitude = 37.754, zoom = 13),
  layers = list(
    list(
      `@@type` = "GeoArrowSolidPolygonLayer",
      data = list(type = "duckdb", query = "SELECT geometry FROM zones"),
      getPolygon = "@@col:geometry",
      getFillColor = c(0, 200, 100, 150),
      getLineColor = c(0, 100, 50),
      lineWidthMinPixels = 2
    )
  )
)

deckgl(spec, data = list(zones = df))
```

**Note**: Use native `POLYGON_2D` type with `ST_MakePolygon()` and `ST_MakeLine()` for proper GeoArrow polygon rendering. The WKB format from `ST_GeomFromText()` has compatibility issues.

## Shiny Integration

Shiny apps use binary Arrow streaming (no Base64 encoding):

```{r shiny, eval=FALSE}
library(shiny)
library(rDeckgl)
library(duckdb)
library(DBI)

ui <- fluidPage(
  sliderInput("n", "Number of points:", 100, 5000, 1000, step = 100),
  deckglOutput("map", height = "600px")
)

server <- function(input, output, session) {
  output$map <- renderDeckgl({
    # Generate spatial points
    con <- dbConnect(duckdb())
    dbExecute(con, "INSTALL spatial; LOAD spatial;")

    dbExecute(con, sprintf("
      CREATE TABLE pts AS
      SELECT
        ST_Point(
          -122.4 + (random() - 0.5) * 0.2,
          37.78 + (random() - 0.5) * 0.15
        )::POINT_2D as geometry
      FROM range(%d)
    ", input$n))

    df <- dbGetQuery(con, "SELECT * FROM pts")
    dbDisconnect(con)

    spec <- list(
      initialViewState = list(longitude = -122.4, latitude = 37.78, zoom = 11),
      layers = list(
        list(
          `@@type` = "GeoArrowScatterplotLayer",
          data = list(type = "duckdb", query = "SELECT geometry FROM pts"),
          getPosition = "@@col:geometry",
          getFillColor = c(255, 100, 0),
          getRadius = 60
        )
      )
    )

    deckgl(spec, data = list(pts = df))
  })
}

shinyApp(ui, server)
```

## Advanced: Custom Bundle Modifications

The `deckgl-bundle.js` includes three key customizations:

1. **WKB→GeoArrow Polygon Converter**: DuckDB's `ST_AsWKB()` outputs binary blobs. GeoArrow layers expect `List<List<FixedSizeList<2, Float64>>>`. The bundle includes a runtime converter that:
   - Parses WKB geometry type and offsets
   - Allocates Arrow buffers for polygon/ring/point hierarchy
   - Constructs typed Arrow vectors with proper metadata

2. **Geometry Metadata Patching**: DuckDB's Arrow export omits `ARROW:extension:name` metadata. The bundle detects binary/list columns named `geometry` and infers types:
   - Binary → `geoarrow.wkb`
   - List → `geoarrow.polygon`

3. **Unified Layer Export**: Merges `@deck.gl/layers`, `@deck.gl/geo-layers`, and `@geoarrow/deck.gl-layers` into single `window.rDeckgl.Layers` namespace for consistent R access.

These modifications enable seamless spatial data visualization without manual geometry conversion in R.

## Further Reading

- [deck.gl Layer Catalog](https://deck.gl/docs/api-reference/layers)
- [DuckDB Spatial Extension](https://duckdb.org/docs/extensions/spatial)
- [GeoArrow Specification](https://github.com/geoarrow/geoarrow)
